graphsim: An R package for simulating gene expression data from graph structures of biological pathways

### 2020-06-24 paper/paper.Rmd
update paper formatting (biorXiv version)

Assuming a unit variance for each gene, these values can be used to
derive a $\Sigma$ matrix: $$\Sigma_{ij} = 
\begin{cases}
   1  & \mbox{if } i=j \\
   \rho R_{ij}  & \mbox{otherwise}
\end{cases}$$ where $\rho$ is the correlation between adjacent nodes.
Thus covariances between adjacent nodes are assigned by a correlation
parameter ($\rho$) and the remaining off-diagonal values in the matrix
are based on scaling these correlations by the geometrically weighted
relationship matrix (or the nearest positive definite matrix for
$\Sigma$ with negative correlations).\
Computing the nearest positive definite matrix is necessary to ensure
that the variance-covariance matrix could be inverted when used as a
parameter in multivariate normal simulations, particularly when negative
correlations are included for inhibitions (as shown below). Matrices
that could not be inverted occurred rarely with biologically plausible
graph structures but this approach allows for the computation of a
plausible correlation matrix when the given graph structure is
incomplete or contains loops. When required, the nearest positive
definite matrix is computed using the `nearPD` function of the \texttt{Matrix} R package
[@Matrix] to perform Higham's algorithm [@Higham2002] on
variance-covariance matrices. The \texttt{graphsim} package gives a warning when this
occurs.

@@

The graph structure in
Figure [1a](#fig:simple_graph:first){reference-type="ref"
reference="fig:simple_graph:first"} was used to simulate correlated gene
expression data by sampling from a multivariate normal distribution
using the R package [@Genz2009; @mvtnorm]. The graph structure
visualisation in
Figure [1](#fig:simple_graph){reference-type="ref"
reference="fig:simple_graph"} was specifically developed for (directed)
\texttt{igraph} objects in and is available in the and packages. The
\texttt{plot_directed} function enables customisation of plot parameters for
each node or edge, and mixed (directed) edge types for indicating
activation or inhibition. These inhibition links (which occur frequently
in biological pathways) are demonstrated in
Figure [1b](#fig:simple_graph:second){reference-type="ref"
reference="fig:simple_graph:second"}.

>>>>

Assuming a unit variance for each gene, these values can be used to
derive a $\Sigma$ matrix: $$\Sigma_{ij} = 
\begin{cases}
   1  & \mbox{if } i=j \\
   \rho R_{ij}  & \mbox{otherwise}
\end{cases}$$ where $\rho$ is the correlation between adjacent nodes.
Thus covariances between adjacent nodes are assigned by a correlation
parameter ($\rho$) and the remaining off-diagonal values in the matrix
are based on scaling these correlations by the geometrically weighted
relationship matrix (or the nearest positive definite matrix for
$\Sigma$ with negative correlations).\

Computing the nearest positive definite matrix is necessary to ensure
that the variance-covariance matrix could be inverted when used as a
parameter in multivariate normal simulations, particularly when negative
correlations are included for inhibitions (as shown below). Matrices
that could not be inverted occurred rarely with biologically plausible
graph structures but this approach allows for the computation of a
plausible correlation matrix when the given graph structure is
incomplete or contains loops. When required, the nearest positive
definite matrix is computed using the `nearPD` function of the \texttt{Matrix} R package
[@Matrix] to perform Higham's algorithm [@Higham2002] on
variance-covariance matrices. The \texttt{graphsim} package gives a warning when this
occurs.

@@

The graph structure in
Figure [1a](#fig:simple_graph:first){reference-type="ref"
reference="fig:simple_graph:first"} was used to simulate correlated gene
expression data by sampling from a multivariate normal distribution
using the R package [@Genz2009; @mvtnorm]. The graph structure
visualisation in
Figure [1](#fig:simple_graph){reference-type="ref"
reference="fig:simple_graph"} was specifically developed for (directed)
\texttt{igraph} objects in and is available in the and packages. The
\texttt{plot\_directed} function enables customisation of plot parameters for
each node or edge, and mixed (directed) edge types for indicating
activation or inhibition. These inhibition links (which occur frequently
in biological pathways) are demonstrated in
Figure [1b](#fig:simple_graph:second){reference-type="ref"
reference="fig:simple_graph:second"}.
